---
title: alm tutorial
layout: tutorial
packge_version: 0.1.9
---

```{r, eval=TRUE, echo=FALSE}
knitr::opts_chunk$set(fig.path="../assets/tutorial-images/alm/")
```

## What are article level metrics?

Glad you asked. The canonical URL for this is perhaps [altmetrics.org](http://altmetrics.org/manifesto/). Basically it is a metric that measures something about an article. This is in stark contrast to journal level metrics, like the Journal Impact Factor.

## Are there other altmetrics data providers?

Yes indeedy.

+ [ImpactStory](http://impactstory.it/)
+ [Altmetric.com](http://altmetric.com/)
+ [PlumAnalytics](http://www.plumanalytics.com/)

## Authentication

You are now required to use an API key to access the PLoS ALM API.

Get your API key [here](http://alm.plos.org/)

Put your API key in your .Rprofile file using exactly this:
options(PlosApiKey = "YOUalmAPIKEY"),
and the functions within this package will be able to use your API key without you having to enter it every time you run a search.

If you want to use Crossref, eLife, or other data providers that have built on top of the original PLOS ALM application, go to their respective web sites to get an API key for their data. E.g., you can't use the PLOS ALM key to access Crossref ALM data.

* Public Knowledg Project (PKP) - http://pkp-alm.lib.sfu.ca/
* Crossref - http://alm.labs.crossref.org/
* eLife - http://alm.svr.elifesciences.org/

<section id="installation">

## Installation

You can get this package by installing via `install_github()` within Hadley Wickham's devtools package.


```{r eval=FALSE}
install.packages("devtools")
require(devtools)
install_github("alm", "rOpenSci")
```



```{r message=FALSE, warning=FALSE}
library('alm')
```

<section id="usage">

## Usage

### The default call with either doi, pmid, pmcid, or mdid without specifying an argument for info

(We'll not print a few columns so the table prints nicely)

```{r alm1, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
alm(doi="10.1371/journal.pone.0029797")[,-c(6:8)]
```


### Details for a single DOI

```{r alm2, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
out <- alm(doi='10.1371/journal.pone.0029797', info='detail')
### totals
out[["totals"]][,-c(6:8)]
```

### Search using various identifiers, including pubmed id, pmc id, and mendeley id

```{r alm3, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
# A single PubMed ID (pmid)
alm(pmid=22590526)[,-c(6:8)]

# A single PubMed Central ID (pmcid)
alm(pmcid=212692)[,-c(6:8)]
```


### Search on many identifiers

```{r xxx, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
dois <- c('10.1371/journal.pone.0001543','10.1371/journal.pone.0040117',
  '10.1371/journal.pone.0029797','10.1371/journal.pone.0039395')
out <- alm(doi=dois)
lapply(out, head)
```

### Get altmetrics by year

You can also get metrics by day (`sum_metrics='day'`) or month (`sum_metrics='month'`)

```{r byyear, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
alm(doi='10.1371/journal.pone.0036240', sum_metrics='year')[,-c(6:8)]
```

### Output an-easy-to-combine-with-other-results data.frame

```{r easytocombine, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
alm(doi='10.1371/journal.pone.0035869', total_details=TRUE)[,3:10]
```

### Get detailed data for altmetrics using `almevents`

```{r almevents, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
out <- almevents(doi="10.1371/journal.pone.0029797")
names(out) # names of sources
out <- out[!out %in% c("sorry, no events content yet","parser not written yet")] # remove those with no data
out[["pmc"]] # get the results for PubMed Central
out[["twitter"]][1:3,] # get the results for twitter
```

### Alt-metrics total citations from all sources.

```{r almtotals, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
almtotals(doi = '10.1371/journal.pbio.0000012')
```

### Get title of article by inputting the doi for the article.

```{r almtitle, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
almtitle(doi='10.1371/journal.pbio.0000012')
```


### Retrieve and plot PLOS article-level metrics signposts.

```{r signposts1, message=FALSE, warning=FALSE, comment=NA, cache=TRUE, fig.keep='high'}
dat <- signposts(doi="10.1371/journal.pone.0029797")
plot_signposts(input=dat)
```

Or plot many identifiers gives a line chart

```{r signposts2, message=FALSE, warning=FALSE, comment=NA, cache=TRUE, fig.width=7}
dois <- c('10.1371/journal.pone.0001543','10.1371/journal.pone.0040117','10.1371/journal.pone.0029797','10.1371/journal.pone.0039395')
dat <- signposts(doi=dois)
plot_signposts(input=dat)
```

Or make an interactive chart by doing `plot_signposts(input=dat, type="multiBarChart")`. Try it out! It should open in your browser and you can interact with it.

### Density and histogram plots from PLOS Article Level Metrics data

Note: Do you the key below in the `searchplos` call in this example, but if you plan to use rplos more, get your own API key [here](http://api.plos.org/).

```{r plot_densityprep, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
library('rplos'); library('plyr')
dois <- searchplos(q='science', fl="id", fq=list('cross_published_journal_key:PLoSONE', 'doc_type:full', 'publication_date:[2010-01-01T00:00:00Z TO 2010-12-31T23:59:59Z]'), limit=200)
```

Remove non-full article DOIs

```{r remove}
dois <- dois$id
dois <- dois[!grepl("annotation", dois)]
```

Collect altmetrics data and combine to a `data.frame` with `ldply`

```{r plot_densityprep2, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
alm <- alm(doi=dois, total_details=TRUE)
alm <- ldply(alm)
```

The default plot

```{r plot_densityplot1, message=FALSE, warning=FALSE, comment=NA, cache=TRUE, fig.width=5, fig.height=4}
plot_density(alm)
```

You can change the color of the density plot

```{r plot_densityplot2, message=FALSE, warning=FALSE, comment=NA, cache=TRUE, fig.width=5, fig.height=4}
plot_density(alm, color="#EFA5A5")
```

Pass in a title or description subtending the title

```{r plot_densityplot3, message=FALSE, warning=FALSE, comment=NA, cache=TRUE, fig.width=5, fig.height=4}
plot_density(alm, title="Scopus citations from 2010")
```

Plot a particular source

```{r plot_densityplot4, message=FALSE, warning=FALSE, comment=NA, cache=TRUE, fig.width=5, fig.height=4}
names(alm)[1:35]
plot_density(input=alm, source="crossref_citations")
```

Plot many sources in different panels in the same plot, and pass in colors just for fun

```{r plot_densityplot5, message=FALSE, warning=FALSE, comment=NA, cache=TRUE, fig.width=7, fig.height=6}
plot_density(input=alm, source=c("counter_total","crossref_citations","twitter_total"), color=c("#83DFB4","#EFA5A5","#B2C9E4"))
```

### CrossRef article level metrics

Remember to get your api key from CrossRef, pass it in in the key parameter. Notice that we are passing the base url for the Crossref API, whereas the default is for the PLOS url [http://alm.plos.org/api/v3/articles](http://alm.plos.org/api/v3/articles).

```{r crossref, message=FALSE, warning=FALSE, comment=NA, cache=TRUE}
url <- "http://alm.labs.crossref.org/api/v3/articles"
dois <- c("10.1371/journal.pone.0086859", "10.1038/nature12990", "10.5860/choice.51-3037")
alm(doi=dois, url = url, key=getOption("crossrefalmkey"))
```

<section id="citing">

## Citing

To cite `alm` in publications use:

<br>

> Scott Chamberlain, Carl Boettiger, Karthik Ram and Fenner Martin (2014). alm: R wrapper to the almetrics API platform developed by PLoS. R package version 0.1.9. https://github.com/ropensci/alm

<section id="license_bugs">

## License and bugs

* License: [CC0](http://creativecommons.org/choose/zero/)
* Report bugs at [our Github repo for alm](https://github.com/ropensci/alm/issues?state=open)

[Back to top](#top)
