---
title: bold tutorial
layout: tutorial
packge_version: 0.2.0
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
	fig.path="../assets/tutorial-images/bold/",
	comment = "#>", 
	collapse = TRUE,
	warning = FALSE, 
	message = FALSE
)
```

`bold` is an R package to connect to [BOLD Systems](http://www.boldsystems.org/) via their API. Functions in `bold` let you search for sequence data, specimen data, sequence + specimen data, and download raw trace files.

__`bold` info__

+ [BOLD home page](http://boldsystems.org/)
+ [BOLD API docs](boldsystems.org/index.php/resources/api)

<section id="installation">

## Installation

You can install the stable version from CRAN


```{r eval=FALSE}
install.packages("bold")
```

Or the development version from Github


```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("ropensci/bold")
```

Then load the package into the R sesssion


```{r}
library("bold")
```

<section id="usage">

## Usage

## Search for taxonomic names via names

`bold_tax_name` searches for names with names.

```{r cache=FALSE}
bold_tax_name(name='Diplura')
```

```{r cache=FALSE}
bold_tax_name(name=c('Diplura','Osmia'))
```


### Search for taxonomic names via BOLD identifiers

`bold_tax_id` searches for names with BOLD identifiers.

```{r cache=FALSE}
bold_tax_id(id=88899)
```

```{r cache=FALSE}
bold_tax_id(id=c(88899,125295))
```


### Search for sequence data only

The BOLD sequence API gives back sequence data, with a bit of metadata.

The default is to get a list back

```{r cache=FALSE}
bold_seq(taxon='Coelioxys')[1:2]
```

You can optionally get back the `httr` response object

```{r cache=FALSE}
res <- bold_seq(taxon='Coelioxys', response=TRUE)
res$headers
```

You can do geographic searches

```{r cache=FALSE}
bold_seq(geo = "USA")
```

And you can search by researcher name

```{r cache=FALSE}
bold_seq(researchers='Thibaud Decaens')[[1]]
```

by taxon IDs

```{r cache=FALSE}
bold_seq(ids=c('ACRJP618-11','ACRJP619-11'))
```

by container (containers include project codes and dataset codes)

```{r cache=FALSE}
bold_seq(container='ACRJP')[[1]]
```

by bin (a bin is a _Barcode Index Number_)

```{r cache=FALSE}
bold_seq(bin='BOLD:AAA5125')[[1]]
```

And there are more ways to query, check out the docs for `?bold_seq`.


### Search for specimen data only

The BOLD specimen API doesn't give back sequences, only specimen data. By default you download `tsv` format data, which is given back to you as a `data.frame`

```{r cache=FALSE}
res <- bold_specimens(taxon='Osmia')
head(res[,1:8])
```

You can optionally get back the data in `XML` format

```{r eval=FALSE}
bold_specimens(taxon='Osmia', format='xml')
```

```{r eval=FALSE, results='asis'}
<?xml version="1.0" encoding="UTF-8"?>
<bold_records  xsi:noNamespaceSchemaLocation="http://www.boldsystems.org/schemas/BOLDPublic_record.xsd"  xmlns:xsi="http://www.w3.org/2001/XMLSchema-instance">
  <record>
    <record_id>1470124</record_id>
    <processid>BOM1525-10</processid>
    <bin_uri>BOLD:AAN3337</bin_uri>
    <specimen_identifiers>
      <sampleid>DHB 1011</sampleid>
      <catalognum>DHB 1011</catalognum>
      <fieldnum>DHB1011</fieldnum>
      <institution_storing>Marjorie Barrick Museum</institution_storing>
    </specimen_identifiers>
    <taxonomy>
```

You can choose to get the `httr` response object back if you'd rather work with the raw data returned from the BOLD API.

```{r}
res <- bold_specimens(taxon='Osmia', format='xml', response=TRUE)
res$url
res$status_code
res$headers
```

### Search for specimen plus sequence data

The specimen/sequence combined API gives back specimen and sequence data. Like the specimen API, this one gives by default `tsv` format data, which is given back to you as a `data.frame`. Here, we're setting `sepfasta=TRUE` so that the sequence data is given back as a list, and taken out of the `data.frame` returned so the `data.frame` is more manageable.

```{r cache=FALSE}
res <- bold_seqspec(taxon='Osmia', sepfasta=TRUE)
res$fasta[1:2]
```

Or you can index to a specific sequence like

```{r cache=FALSE}
res$fasta['GBAH0293-06']
```

### Get trace files

This function downloads files to your machine - it does not load them into your R session - but prints out where the files are for your information.

```{r eval=FALSE}
bold_trace(taxon='Osmia', quiet=TRUE)
#> Downloading: 51 MB     
#> <bold trace files> 
#> 
#> .../bold/bold_trace_files/BBHYL361-10[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/BBHYL361-10[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/BBHYL363-10[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/BBHYL363-10[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/BBHYL365-10[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/BBHYL365-10[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBAPB666-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBAPB666-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBAPB667-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBAPB674-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBAPB674-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBAPB685-09[LepF1,C_ANTMR1D]_F.ab1
#> .../bold/bold_trace_files/FBAPB685-09[RonMWASPdeg_t1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBAPB685-09[RonMWASPdeg_t1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBAPB708-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBAPB708-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBAPB718-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBAPB718-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBAPB739-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBAPB739-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBHAP295-09[LepF1,C_ANTMR1D]_F.ab1
#> .../bold/bold_trace_files/FBHAP295-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBHAP295-09[RonMWASPdeg_t1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBHAP295-09[RonMWASPdeg_t1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBHAP299-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBHAP299-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBHAP301-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBHAP301-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBHAP328-09[LepF1,C_ANTMR1D]_F.ab1
#> .../bold/bold_trace_files/FBHAP328-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBHAP328-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBHAP328-09[RonMWASPdeg_t1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBHAP328-09[RonMWASPdeg_t1,LepR1]_R.ab1
#> .../bold/bold_trace_files/FBHAP329-09[LepF1,LepR1]_F.ab1
#> .../bold/bold_trace_files/FBHAP329-09[LepF1,LepR1]_R.ab1
#> .../bold/bold_trace_files/SSWLB1482-13[LepF1,LepR1]_R.ab1
```


<section id="citing">

## Citing

To cite `bold` in publications use:

<br>

> Scott Chamberlain (2014). bold: Interface to Bold Systems API. R package version 0.2.0. https://github.com/ropensci/bold

<section id="license_bugs">

## License and bugs

* License: [MIT](http://opensource.org/licenses/MIT)
* Report bugs at [our Github repo for bold](https://github.com/ropensci/bold/issues?state=open)

[Back to top](#top)
