---
title: rfisheries tutorial
layout: tutorial
packge_version: 0.1
---

```{r, eval=TRUE, echo=FALSE}
opts_chunk$set(fig.path="../assets/tutorial-images/rfisheries/")
```

This package provides programmatic access to the [openfisheries](http://openfisheries.org/) [API](http://openfisheries.org/api-info).

Open Fisheries is a platform that aggregates global fishery data and currently offers global fish capture landings from 1950 onwards (more data coming soon). Read more about that effort [here](http://openfisheries.org/about).

<section id="installation">

## Installation


```{r install, eval=FALSE}
install.packages("rfisheries")
```

<section id="usage">

## Usage

Package currently provides three basic functions. Landings data can be obtained by calling `landings()`

```{r landings, messages=FALSE, warning=FALSE}
library(rfisheries)
head( of_landings() )
```

To get country specific data, provide a iso3c country code

```{r landings_co, messages=FALSE, warning=FALSE}
head( of_landings(country = "USA") )
```

To get species specific landings, provide the correct a3 code for the required species.

```{r landings_sp, messages=FALSE, warning=FALSE}
head( of_landings(species = "SKJ") )
```

If you don't have know the correct species or country codes, you can get a complete list with the following two functions.

```r
head( species_codes() )

         scientific_name   taxocode a3_code isscaap
1     Petromyzon marinus 1020100101     LAU      25
2   Lampetra fluviatilis 1020100201     LAR      25
3    Lampetra tridentata 1020100202     LAO      25
4 Ichthyomyzon unicuspis 1020100401     LAY      25
5    Eudontomyzon mariae 1020100501     LAF      25
6      Geotria australis 1020100701     LAE      25
             english_name
1             Sea lamprey
2           River lamprey
3         Pacific lamprey
4          Silver lamprey
5 Ukrainian brook lamprey
6         Pouched lamprey
```

Similarly you can get a full list of country codes

```{r countrycodes, messages=FALSE, warning=FALSE}
head( of_country_codes() )
```

Example: Compare landings from multiple countries

```{r getdata, messages=FALSE, warning=FALSE, tidy=FALSE}
library(plyr)
countries <- of_country_codes()
# let's take a small subset, say 5 random countries
c_list <- countries[sample(nrow(countries), 5),]$iso3c
# and grab landings data for these countries
results <- ldply(c_list, function(x) {
    df <- of_landings(country = x)
    df$country  <-  x
    df
})
```

You can easily compare these results

```{r plot, messages=FALSE, warning=FALSE, tidy=FALSE}
library(ggplot2)
ggplot(results, aes(year, catch, group = country, color = country)) +
  geom_line()
```

Similarly you can get landings data for multiple species. As the API evolves, we'll update the package.


<section id="citing">

## Citing

To cite `rfisheries` in publications use:

<br>

> Karthik Ram, Carl Boettiger and Andrew Dyck (2013). rfisheries: R interface for fisheries data. R package version 0.1. http://www.github.com/ropensci/rfisheries

<section id="license_bugs">

## License and bugs

* License: [CC0](http://creativecommons.org/choose/zero/)
* Report bugs at [our Github repo for rfisheries](https://github.com/ropensci/rfisheries/issues?state=open)
