---
title: rgauges tutorial
layout: tutorial
packge_version: 0.2.0
---

```{r, eval=TRUE, echo=FALSE}
opts_chunk$set(fig.path="../assets/tutorial-images/rgauges/")
```

`rgauges` is an R wrapper to the gaug.es API for website analytics.

### Gaug.es API docs

See the Gaug.es API documentation here [http://get.gaug.es/documentation/](http://get.gaug.es/documentation/)

### Authentication

Get your own API key in your Gaug.es 'My Account' page and put in your .Rprofile file under the name 'GaugesKey' or some other name (you can specify `keyname` in function calls - but if you put in as 'GaugesKey' you're all set and don't need to bother with the `keyname` parameter).


<section id="installation">

## Installation


More stable version from CRAN

```{r installcran, messages=FALSE, warnings=FALSE, eval=FALSE}
install.packages("rgauges")
```

Development version from Github

```{r installgithub, messages=FALSE, warnings=FALSE, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("rgauges", "ropensci")
```

Load

```{r load, messages=FALSE, warnings=FALSE, cache=FALSE}
library(rgauges)
```

<section id="usage">

## Usage


#### Your info

```{r yourinfo, messages=FALSE, warnings=FALSE, cache=FALSE}
gs_me()
```

##### Traffic

```{r traffic, messages=FALSE, warnings=FALSE, cache=FALSE}
gs_traffic(id='4efd83a6f5a1f5158a000004')
```

#### Screen/browser information

```{r screenbrowser, messages=FALSE, warnings=FALSE, cache=FALSE}
gs_reso(id='4efd83a6f5a1f5158a000004')
```

#### Visualize traffic data

You'll need to load ggplot2

```{r viz, messages=FALSE, warnings=FALSE, cache=FALSE}
library(ggplot2)
out <- gs_gauge_detail(id='4efd83a6f5a1f5158a000004')
vis_gauge(out)
```

<section id="citing">

## Citing

To cite `rgauges` in publications use:

<br>

> Scott Chamberlain and Karthik Ram (2014). rgauges: R wrapper to Gaug.es API. R package version 0.2.0. https://github.com/ropensci/rgauges

<section id="license_bugs">

## License and bugs

* License: [MIT](http://opensource.org/licenses/MIT)
* Report bugs at [our Github repo for rgauges](https://github.com/ropensci/rgauges/issues?state=open)

[Back to top](#top)
