---
title: rgbif tutorial
layout: tutorial
packge_version: 0.6.2
---

```{r, eval=TRUE, echo=FALSE}
opts_chunk$set(fig.path="../assets/tutorial-images/rgbif/", fig.width=6, fig.height=4)
```

`rgbif` is an R package to search and retrieve data from the Global Biodiverity Information Facilty (GBIF). `rgbif` wraps R code around the [GBIF API][gbifapi] to allow you to talk to GBIF from R.

********************

<section id="installation">

## Installation

Install rgbif

```{r install, eval=FALSE}
install.packages("rgbif")
```

Load rgbif

```{r load, comment=NA, warning=FALSE, message=FALSE}
library("rgbif")
```

********************

<section id="usage">

## Usage

### Get number of occurrences for a set of search parameters

Search by type of record, all observational in this case

```{r occ_count1, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_count(basisOfRecord='OBSERVATION')
```

Records for **Puma concolor** with lat/long data (georeferened) only

```{r occ_count2, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_count(taxonKey=2435099, georeferenced=TRUE)
```

All georeferenced records in GBIF

```{r occ_count3, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_count(georeferenced=TRUE)
```

Records from Denmark

```{r occ_count4, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_count(country='DENMARK')
```

Number of records in a particular dataset

```{r occ_count6, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_count(datasetKey='9e7ea106-0bf8-4087-bb61-dfe4f29e0f17')
```

All records from 2012

```{r occ_count7, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_count(year=2012)
```

********************

### Get possible values to be used in taxonomic rank arguments in functions

```{r taxrank, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
taxrank()
```

********************

### Search for taxon information

Search for a genus

```{r name_lookup1, comment=NA, warning=FALSE, message=FALSE}
head(name_lookup(query='Cnaemidophorus', rank="genus", return="data"))
```

Search for the class mammalia

```{r name_lookup2, comment=NA, warning=FALSE, message=FALSE}
head(name_lookup(query='mammalia')$data)
```

Look up the species Helianthus annuus

```{r name_lookup3, comment=NA, warning=FALSE, message=FALSE}
head(name_lookup('Helianthus annuus', rank="species")$data)
```

********************

### Get data for a single occurrence. Note that data is returned as a list, with slots for metadata and data, or as a hierarchy, or just data.

Just data

```{r occ_get1, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_get(key=855998194, return='data')
```

Just taxonomic hierarchy

```{r occ_get2, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_get(key=855998194, return='hier')
```

All data, or leave return parameter blank

```{r occ_get3, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_get(key=855998194, return='all')
```

Get many occurrences. `occ_get` is vectorized

```{r occ_get4, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
occ_get(key=c(101010,855998194,49819470), return='data')
```

********************

### Maps

Static map using the ggplot2 package

Make a map of **Puma concolor** occurrences

```{r gbifmap1, comment=NA, warning=FALSE, message=FALSE, cache=FALSE}
key <- name_backbone(name='Puma concolor', kingdom='plants')$speciesKey
dat <- occ_search(taxonKey=key, return='data', limit=300)
dat$name <- "Puma concolor" # so that we just get one symbol type
gbifmap(input=dat)
```

[gbifapi]: http://www.gbif.org/developer/summary


<section id="citing">

## Citing

To cite `rgbif` in publications use:

<br>

> Scott Chamberlain, Carl Boettiger, Karthik Ram, Vijay Barve and Dan Mcglinn (2014). rgbif: Interface to the Global Biodiversity Information Facility API. R package version 0.6.3. https://github.com/ropensci/rgbif

<section id="license_bugs">

## License and bugs

* License: [CC0](http://creativecommons.org/choose/zero/)
* Report bugs at [our Github repo for rgbif](https://github.com/ropensci/rgbif/issues?state=open)

[Back to top](#top)
