---
title: rgbif tutorial
layout: tutorial
packge_version: 0.8.6
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
	fig.path="../assets/tutorial-images/rgbif/",
	fig.width=6,
	fig.height=4,
	comment = "#>",
	collapse = TRUE,
	warning = FALSE,
	message = FALSE
)
```

`rgbif` is an R package to search and retrieve data from the Global Biodiverity Information Facilty (GBIF). `rgbif` wraps R code around the [GBIF API][gbifapi] to allow you to talk to GBIF from R.

********************

<section id="installation">

## Installation

Install `rgbif` from CRAN for more stable version

```{r eval=FALSE}
install.packages("rgbif")
```

Or from Github for the development version

```{r eval=FALSE}
devtools::install_github("ropensci/rgbif")
```

Load rgbif

```{r load}
library("rgbif")
```

********************

<section id="usage">

## Usage

## Number of occurrences

Search by type of record, all observational in this case

```{r occ_count1}
occ_count(basisOfRecord='OBSERVATION')
```

Records for **Puma concolor** with lat/long data (georeferened) only. Note that `hasCoordinate` in `occ_search()` is the same as `georeferenced` in `occ_count()`.

```{r occ_count2}
occ_count(taxonKey=2435099, georeferenced=TRUE)
```

All georeferenced records in GBIF

```{r occ_count3}
occ_count(georeferenced=TRUE)
```

Records from Denmark

```{r occ_count4}
denmark_code <- isocodes[grep("Denmark", isocodes$name), "code"]
occ_count(country=denmark_code)
```

Number of records in a particular dataset

```{r occ_count6}
occ_count(datasetKey='9e7ea106-0bf8-4087-bb61-dfe4f29e0f17')
```

All records from 2012

```{r occ_count7}
occ_count(year=2012)
```

Records for a particular dataset, and only for preserved specimens

```{r occ_count8}
occ_count(datasetKey='e707e6da-e143-445d-b41d-529c4a777e8b', basisOfRecord='OBSERVATION')
```

## Search for taxon names

Get possible values to be used in taxonomic rank arguments in functions

```{r taxrank}
taxrank()
```

`name_lookup()` does full text search of name usages covering the scientific and vernacular name, the species description, distribution and the entire classification across all name usages of all or some checklists. Results are ordered by relevance as this search usually returns a lot of results.

By default `name_lookup()` returns five slots of information: meta, data, facets, hierarchies, and names. hierarchies and names elements are named by their matching GBIF key in the `data.frame` in the data slot.

```{r}
out <- name_lookup(query='mammalia')
```

```{r}
names(out)
```

```{r}
out$meta
```

```{r}
head(out$data)
```

```{r}
out$facets
```

```{r}
out$hierarchies[1:2]
```

```{r}
out$names[2]
```

Search for a genus

```{r name_lookup1}
head(name_lookup(query='Cnaemidophorus', rank="genus", return="data"))
```

Search for the class mammalia

```{r name_lookup2}
head(name_lookup(query='mammalia', return = 'data'))
```

Look up the species Helianthus annuus

```{r name_lookup3}
head(name_lookup('Helianthus annuus', rank="species", return = 'data'))
```

The function `name_usage()` works with lots of different name endpoints in GBIF, listed at [http://www.gbif.org/developer/species#nameUsages]().

```{r}
library("plyr")
out <- name_usage(key=3119195, language="FRENCH", data='vernacularNames')
head(out$data)
```

The function `name_backbone()` is used to search against the GBIF backbone taxonomy

```{r}
name_backbone(name='Helianthus', rank='genus', kingdom='plants')
```

The function `name_suggest()` is optimized for speed, and gives back suggested names based on query parameters.

```{r}
head( name_suggest(q='Puma concolor') )
```


## Single occurrence records

Get data for a single occurrence. Note that data is returned as a list, with slots for metadata and data, or as a hierarchy, or just data.

Just data

```{r occ_get1}
occ_get(key=766766824, return='data')
```

Just taxonomic hierarchy

```{r occ_get2}
occ_get(key=766766824, return='hier')
```

All data, or leave return parameter blank

```{r occ_get3}
occ_get(key=766766824, return='all')
```

Get many occurrences. `occ_get` is vectorized

```{r occ_get4}
occ_get(key=c(766766824, 101010, 240713150, 855998194), return='data')
```


## Search for occurrences

By default `occ_search()` returns a `dplyr` like output summary in which the data printed expands based on how much data is returned, and the size of your window. You can search by scientific name:

```{r}
occ_search(scientificName = "Ursus americanus", limit = 20)
```

Or to be more precise, you can search for names first, make sure you have the right name, then pass the GBIF key to the `occ_search()` function:

```{r}
key <- name_suggest(q='Helianthus annuus', rank='species')$key[1]
occ_search(taxonKey=key, limit=20)
```

Like many functions in `rgbif`, you can choose what to return with the `return` parameter, here, just returning the metadata:

```{r}
occ_search(taxonKey=key, return='meta')
```

You can choose what fields to return. This isn't passed on to the API query to GBIF as they don't allow that, but we filter out the columns before we give the data back to you.

```{r}
occ_search(scientificName = "Ursus americanus", fields=c('name','basisOfRecord','protocol'), limit = 20)
```

Most parameters are vectorized, so you can pass in more than one value:

```{r}
splist <- c('Cyanocitta stelleri', 'Junco hyemalis', 'Aix sponsa')
keys <- sapply(splist, function(x) name_suggest(x)$key[1], USE.NAMES=FALSE)
occ_search(taxonKey=keys, limit=5)
```

## Downloads

GBIF provides access to their downloads API. This is the same service that you use when you go to [their website]() to search for records then download a zip file.

First, we can search for data, for _Ursus arctos_ (`taxonkey` = 2433433)

```{r eval=FALSE}
(res <- occ_download('taxonKey = 2433433'))
```

Then we can check on the download progress

```{r eval=FALSE}
occ_download_meta(res)
#> <<gbif download metadata>>
#>   Status: RUNNING
#> Download key: 0007658-150615163101818
#>   Created: 2015-07-09T14:48:40.705+0000
#>   Modified: 2015-07-09T14:48:40.788+0000
#>   Download link: http://api.gbif.org/v1/occurrence/download/request/0007658-150615163101818.zip
#>   Total records: 0
#>   Request:
#>     type:  equals
#>     predicates:
```

We can see that status is `Running`, so we wait a while longer. And when it's ready, download the data

```{r eval=FALSE}
res <- occ_download_get(res, overwrite = TRUE)
#> <<gbif downloaded get>>
#>   Path: ./0007658-150615163101818.zip
#>   File size: 3.1 MB
```

And then import into R, and manipulate

```{r eval=FALSE}
suppressPackageStartupMessages(library("dplyr"))
df <- tbl_df(occ_download_import(res))
df %>%
  filter(nchar(lifeStage) != 0, !is.na(decimalLatitude)) %>%
  select(decimalLatitude, decimalLongitude, lifeStage)

#> Source: local data frame [53 x 3]
#>
#>    decimalLatitude decimalLongitude lifeStage
#> 1         44.65373       -110.54160  JUVENILE
#> 2         63.18333       -139.60000     ADULT
#> 3         70.48056       -157.41667     ADULT
#> 4         52.66667       -117.86667     ADULT
#> 5         59.48333       -139.81667     ADULT
#> 6         69.93456       -148.68292     ADULT
#> 7         69.01667       -161.93333  JUVENILE
#> 8         59.89000       -151.68167     ADULT
#> 9         59.64154       -139.40070     ADULT
#> 10        44.61667        -79.41667     ADULT
#> ..             ...              ...       ...
```

********************

## Maps

Static map using the ggplot2 package. Make a map of *Puma concolor* occurrences.

```{r gbifmap1}
key <- name_backbone(name='Puma concolor')$speciesKey
dat <- occ_search(taxonKey=key, return='data', limit=300)
gbifmap(input=dat)
```

[gbifapi]: http://www.gbif.org/developer/summary


<section id="citing">

## Citing

To cite `rgbif` in publications use:

<br>

> Scott Chamberlain, Carl Boettiger, Karthik Ram, Vijay Barve and Dan Mcglinn (2015). rgbif: Interface to the Global Biodiversity Information Facility API. R package version 0.8.6 https://github.com/ropensci/rgbif

<section id="license_bugs">

## License and bugs

* License: [MIT](http://opensource.org/licenses/MIT)
* Report bugs at [our Github repo for rgbif](https://github.com/ropensci/rgbif/issues?state=open)

[Back to top](#top)
