---
title: rnoaa tutorial
layout: tutorial
packge_version: 0.1.0
---

```{r, eval=TRUE, echo=FALSE}
opts_chunk$set(fig.path="../assets/tutorial-images/rnoaa/", fig.width=6, fig.height=4)
```

<section id="installation">

## Installation


Install and load `rnoaa` into the R session.

```{r install, eval=FALSE}
library("devtools")
devtools::install_github("rnoaa", "ropensci")
```

```{r load, message=FALSE, warning=FALSE}
library("rnoaa")
library("plyr")
```
<br>

<section id="usage">

## Usage

Get info on a station by specifying a datasetid, locationid, and stationid

```{r two, message=FALSE, warning=FALSE, cache=FALSE}
noaa_stations(datasetid='GHCND', locationid='FIPS:12017', stationid='GHCND:USC00084289')
```
<br>
Search for data and get a data.frame or list

```{r three, message=FALSE, warning=FALSE, cache=FALSE}
out <- noaa(datasetid='GHCND', stationid='GHCND:USW00014895', startdate = '2013-10-01', enddate = '2013-12-01')
```

See a data.frame

```{r four, message=FALSE, warning=FALSE, cache=FALSE}
out$data
```


Plotting

Example 1

Search for data first, then plot

```{r}
out <- noaa(datasetid='GHCND', stationid='GHCND:USW00014895', datatypeid='PRCP', startdate = '2010-05-01', enddate = '2010-10-31', limit=500)
```

Default plot

```{r}
noaa_plot(out)
```

Create 14 day breaks

```{r}
noaa_plot(out, breaks="14 days")
```

One month breaks

```{r}
noaa_plot(out, breaks="1 month", dateformat="%d/%m")
```
<br>
Example 2

Search for data

```{r}
out2 <- noaa(datasetid='GHCND', stationid='GHCND:USW00014895', datatypeid='PRCP', startdate = '2010-05-01', enddate = '2010-05-03', limit=100)
```

Make a plot, with 6 hour breaks, and date format with only hour

```{r}
noaa_plot(out2, breaks="6 hours", dateformat="%H")
```
<br>
Combine many calls to noaa function

Search for two sets of data

```{r}
out1 <- noaa(datasetid='GHCND', stationid='GHCND:USW00014895', datatypeid='PRCP', startdate = '2010-03-01', enddate = '2010-05-31', limit=500)

out2 <- noaa(datasetid='GHCND', stationid='GHCND:USW00014895', datatypeid='PRCP', startdate = '2010-09-01', enddate = '2010-10-31', limit=500)
```

Then combine with a call to `noaa_combine`

```{r}
df <- noaa_combine(out1, out2)
head(df[[1]]); tail(df[[1]])
```

Then plot - the default passing in the combined plot plots the data together. In this case it looks kind of weird since a straight line combines two distant dates.

```{r}
noaa_plot(df)
```

But we can pass in each separately, which uses `facet_wrap` in `ggplot2` to plot each set of data in its own panel.

```{r}
noaa_plot(out1, out2, breaks="60 days", dateformat = "%d/%m")
```

<section id="citing">

## Citing

To cite `rnoaa` in publications use:

<br>

> Hart Edmund, Scott Chamberlain and Karthik Ram (2014). rnoaa: NOAA climate data from R. R package version 0.1.2. https://github.com/ropensci/rnoaa

<section id="license_bugs">

## License and bugs

* License: [MIT](http://opensource.org/licenses/MIT)
* Report bugs at [our Github repo for rnoaa](https://github.com/ropensci/rnoaa/issues?state=open)

[Back to top](#top)
