---
name: taxize_v03
layout: post
title: taxize v0.3.0 update - a new data source, taxonomy in writing, and uBio examples
date: 2014-05-20
author: Scott Chamberlain
tags:
- R
- taxonomy
- API
- taxize
---

We just released `v0.3` of `taxize`. For details on the update, see the [release notes](https://github.com/ropensci/taxize/releases/tag/v0.3.0). 

## Some new features

* New function `iplant_resolve()` to do name resolution using the iPlant name resolution service. Note, this is different from http://taxosaurus.org/ that is wrapped in the `tnrs()` function.
* New function `ipni_search()` to search for names in the International Plant Names Index (IPNI). See below for more.
* New function `resolve()` that unifies name resolution services from iPlant's name resolution service (via `iplant_resolve()`), Taxosaurus' TNRS (via `tnrs()`), and GNR's name resolution service (via `gnr_resolve()`).
* All `get_` functions now returning a new _uri_ attribute that is a link to the taxon on the web. If NA is given back (e.g. nothing found), the uri attribute is blank. You can go directly to the uri in your default browser by doing, for example: `browseURL(attr(result, "uri"))`.

## Updating to v0.3

Since `taxize` is not updated to `v0.3` on CRAN yet at the time of writing this, install `taxize` from GitHub:

```{r eval=FALSE}
devtools::install_github("ropensci/taxize")
```

Then load `taxize`

```{r message=FALSE}
library("taxize")
```


## International Plant Names Index (IPNI)

We added the IPNI as a new data source in `taxize` in `v0.3`. Currently, there is only one function to interact with IPNI: `ipni_search()`. What follows are a few examples of how you can use `ipni_search()`.

Search for the genus _Brintonia_

```{r}
ipni_search(genus='Brintonia')[,c(1:3)]
```

Search for the species _Pinus contorta_

```{r}
head(ipni_search(genus='Pinus', species='contorta')[,c(1:3)])
```

Different output formats (the default is _minimal_)

```{r}
head(ipni_search(genus='Ceanothus')[,c(1:3)])
```

```{r}
head(ipni_search(genus='Ceanothus', output='extended'))[,c(1:3)]
```

If you do something wrong, you get a message, and the actual output is `NA`

```{r}
ipni_search(genus='Brintoniaasasf')
```

## uBio examples

Until now, we have had functions to interact with uBio's API, but it probably hasn't been too clear how to use them, and they were a little buggy for sure. We have squashed many bugs in ubio functions. Here is an example workflow of how to use ubio functions.

### ubio_search

Search uBio by taxonomic name. This is sort of the entry point for uBio where you can search by taxonomic name, from which  you can get namebankID's that can be passed to the `ubio_classification_search` and `ubio_namebankID` functions


```{r}
lapply(ubio_search(searchName = 'elephant'), head)
```

```{r}
id <- ubio_search(searchName = 'elephant')$scientific$namebankid[1]
```

### ubio_id

Get data on a specific uBio `namebankID`. Use the id from the previous code block

```{r}
ubio_id(namebankID = id)
```

### ubio_classification_search

Return `hierarchiesID` that refer to the given `namebankID`

```{r}
ubio_classification_search(namebankID = 3070378)
```

### ubio_classification

Return all ClassificationBank data pertaining to a particular `hierarchiesID`

```{r}
ubio_classification(hierarchiesID = 2483153)
```

### ubio_synonyms

Search for taxonomic synonyms by `hierarchiesID`

```{r}
ubio_synonyms(hierarchiesID = 4091702)
```


## Examples of using taxize in writing

Let's say one is writing a paragraph in which you are using taxonomic or common names, and perhaps you want to have the number of taxa in a particular group. You can write a paragaph like:

```{r eval=FALSE}
I studied the common weed species _Tragopogon dubius_ (`r sci2comm('Tragopogon dubius', db='itis')[[1]][1]`; `r tax_name(query = "Tragopogon dubius", get = "family", db = "ncbi")[[1]]`) and _Cirsium arvense_ (`r sci2comm('Cirsium arvense', db='itis')[[1]][1]`; `r tax_name(query = "Cirsium arvense", get = "family", db = "ncbi")[[1]]`).
```

Which renders to:

I studied the common weed species _Tragopogon dubius_ (`r sci2comm('Tragopogon dubius', db='itis')[[1]][1]`; `r tax_name(query = "Tragopogon dubius", get = "family", db = "ncbi")[[1]]`) and _Cirsium arvense_ (`r sci2comm('Cirsium arvense', db='itis')[[1]][1]`; `r tax_name(query = "Cirsium arvense", get = "family", db = "ncbi")[[1]]`).

Notice how inside backticks you can execute code by starting with an _r_, then doing something like searching for common names for a taxon.

### Another example:

```{r eval=FALSE}
We found that `r sci2comm('Tragopogon dubius', db='itis')[[1]][1]` was very invasive.
```

Renders to:

We found that `r sci2comm('Tragopogon dubius', db='itis')[[1]][1]` was very invasive.

### Another example:

```{r eval=FALSE}
There are `r nrow(downstream('Tragopogon', db = "col", downto = "Species")$Tragopogon)` species (source: Catalogue of Life) in the _Tragopogon_ genus, meaning there is much more to study :)
```

Renders to:

There are `r nrow(downstream('Tragopogon', db = "col", downto = "Species")$Tragopogon)` species (source: Catalogue of Life) in the _Tragopogon_ genus, meaning there is much more to study :)


## el fin

Please do update to `v0.3`, try it out, report bugs, and get back to us with any questions!