---
name: solr
layout: post
title: solr - an R interface to Solr
date: 2014-01-27
authors:
  - name: Scott Chamberlain
tags:
- R
- text-mining
- API
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE, 
  message = FALSE
)
```

A number of the APIs we interact with (e.g., PLOS full text API, and USGS's BISON API in [rplos](http://cran.r-project.org/web/packages/rplos/index.html) and [rbison](http://cran.r-project.org/web/packages/rbison/index.html), respectively) expose [Solr][solrweb] endpoints. [Solr][solrweb] is an Apache hosted project - it is a powerful search server.  Given that at least two, and possibly more in the future, of the data providers we interact with provide Solr endpoints, it made sense to create an R package to make robust functions to interact with Solr that work across any Solr endpoint. This is then useful to us, and hopefully others.

The following are a few examples covering some of things you can do in Solr that fall in to six categories:

* Search: via `solr_search`
* Grouping: via `solr_group`
* Faceting: via `solr_facet`
* Highlighting: via `solr_highlight`
* Stats: via `solr_stats`
* More like this: via `solr_mlt`

The `solr` package generally has two steps for any query: a) send the request given your inputs, and b) parse the output into a useful R data structure. Part a) is quite easy. However, part b) is harder. We are working hard on making parsers that are as general as possible for each of the data formats that are returned by group, facet, highlight, etc., but of course we will still definitely fail in many cases. Please do submit bug reports to [our issue tracker](https://github.com/ropensci/solr/issues?state=open) so we can make the parsers work better.

******************

## Installation

`solr` is on CRAN, so you can install the more stable version there, and some dependencies.

```{r eval=FALSE}
install.packages("solr")
```

You can install the development version from Github as follows.  Below we'll use the Github version - most of below is available in the CRAN version too, except `solr_group`.

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("ropensci/solr")
```

Load the library

```{r load}
library("solr")
```

******************

## Define url endpoint and key

As `solr` is a general interface to Solr endpoints, you need to define the url. Here, we'll work with the Public Library of Science full text search API (docs [here](http://api.plos.org/)). Some Solr endpoints will require authentication - I should note that we don't yet handle authentication schemes other than passing in a key in the url, but that's on the to do list.

```{r}
url <- 'http://api.plos.org/search'
```

******************

## Search

```{r}
solr_search(q='*:*', rows=2, fl='id', base=url)
```

Search for words "sports" and "alcohol" within seven words of each other

```{r}
solr_search(q='everything:"sports alcohol"~7', fl='title', rows=3, base=url)
```

******************

## Groups

Most recent publication by journal

```{r}
solr_group(q='*:*', group.field='journal', rows=5, group.limit=1, group.sort='publication_date desc', fl='publication_date, score', base=url)
```

First publication by journal

```{r}
solr_group(q='*:*', group.field='journal', group.limit=1, group.sort='publication_date asc', fl='publication_date, score', fq="publication_date:[1900-01-01T00:00:00Z TO *]", base=url)
```

******************

## Facet

```{r}
solr_facet(q='*:*', facet.field='journal', facet.query='cell,bird', base=url)
```

Range faceting with > 1 field

```{r}
head( solr_facet(q='*:*', base=url, facet.range='alm_twitterCount', facet.range.start=5, facet.range.end=1000, facet.range.gap=10)$facet_ranges$alm_twitterCount )
```

******************

## Highlight

```{r}
solr_highlight(q='alcohol', hl.fl = 'abstract', rows=2, base = url)
```

******************

## Stats

```{r}
solr_stats(q='ecology', stats.field='alm_twitterCount', stats.facet=c('journal','volume'), base=url)
```

******************

## More like this

`solr_mlt` is a function to return similar documents to the ones searched for.

```{r}
out <- solr_mlt(q='title:"ecology" AND body:"cell"', mlt.fl='title', mlt.mindf=1, mlt.mintf=1, fl='counter_total_all', rows=5, base=url)
out$docs
```

******************

## Raw data?

You can optionally get back raw `json` or `xml` from all functions by setting parameter `raw=TRUE`. You can then parse after the fact with `solr_parse`, or just process as you wish. For example:

```{r}
(out <- solr_highlight(q='alcohol', hl.fl = 'abstract', rows=2, base = url, raw=TRUE))
```

Then parse

```{r}
solr_parse(out, 'df')
```

******************

## Verbosity

As you have noticed, we include in each function the acutal call to the Solr endpoint made so you know exactly what was submitted to the remote or local Solr instance. You can suppress the message with `verbose=FALSE`. This message isn't in the CRAN version.

******************

## Advanced: Function Queries

Function Queries allow you to query on actual numeric fields in the SOLR database, and do addition, multiplication, etc on one or many fields to stort results. For example, here, we search on the product of counter_total_all and alm_twitterCount, using a new temporary field "_val_"

```{r}
solr_search(q='_val_:"product(counter_total_all,alm_twitterCount)"', rows=5, fl='id,title', fq='doc_type:full', base=url)
```

Here, we search for the papers with the most citations

```{r}
solr_search(q='_val_:"max(counter_total_all)"', rows=5, fl='id,counter_total_all', fq='doc_type:full', base=url)
```

Or with the most tweets

```{r}
solr_search(q='_val_:"max(alm_twitterCount)"', rows=5, fl='id,alm_twitterCount', fq='doc_type:full', base=url)
```

******************

## Further reading on Solr

+ [Solr home page](http://lucene.apache.org/solr/)
+ [Highlighting help](http://wiki.apache.org/solr/HighlightingParameters)
+ [Faceting help](http://wiki.apache.org/solr/SimpleFacetParameters)
+ [Solr stats](http://wiki.apache.org/solr/StatsComponent)
+ ['More like this' searches](http://wiki.apache.org/solr/MoreLikeThis)
+ [Grouping/Feild collapsing](http://wiki.apache.org/solr/FieldCollapsing)
+ [Installing Solr on Mac using homebrew](http://ramlev.dk/blog/2012/06/02/install-apache-solr-on-your-mac/)
+ [Install and Setup SOLR in OSX, including running Solr](http://risnandar.wordpress.com/2013/09/08/how-to-install-and-setup-apache-lucene-solr-in-osx/)


[solrweb]: http://lucene.apache.org/solr/
