---
name: 2014-in-review
layout: post
title: 2014 in review
date: 2015-01-01
authors:
  - name: Leadership Team
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../assets/blog-images/2015-01-01-2014-in-review/"
)
```

## Software Developed

We actively developed 103 packages (60 of them new to the grant period) over the last 15 month period. The following is a summary of the highlights:

**EML** Ecological Metadata Languages is an XML extension that allows users to make machine readable
metadata for datasets. The R client `EML` makes it simple to create EML formatted metadata, which
can be published to many data repositories online, and used to parse EML data.

**RNeXML** NeXML is a XML extension for specifying metadata, and data, for phylogenetic trees.
Phylogenies are used across many disciplines, including biology, linguistics, and anthropology.
The `RNeXML` R client makes it easy to create and read NeXML.

**spocc** There are a lot of sources of species occurrence data for biologists. We make R clients
for many of them. Instead of users learning a different interface to each data source, `spocc`
allows users to learn one interface to all data sources. We will integrate `rMaps` soon
for interactive maps (see below).

**testdat** Much like `testthat` makes testing easier in R, `testdat` makes cleaning data easier
in R.

**plotly** We accepted into our suite the R client for Plotly, a company that provides an API to
create interactive plots on their web platform.

**rDat** We started development on an R client for Dat, the _github for data_, which will allow our
users to up their reproducible science game by keeping track of changes in their data, and easily
accessing data avaialble in Data repositories.

**rMaps** We are close to accepting rMaps into our software suite, which will be easily incorporated
into especially our package for spatial species occurrence data, `spocc`, but other packages as
well that deal with any kind of spatial data.

**fulltext** A package started in May 2014 aims to be a one stop shop for acquiring data for text
mining in R. In addition to our existing packages to feed into this package for sources of text data,
including for Public Library of Science, Pubmed, eLife, we've started more: for IEEE, arXiv,
Biomed Central, Zenodo, and Crossref.

We now have a number of history R clients, for getting historical USA maps (`USAboundaries`), fetching commonly used history datasets (`historydata`), and determining gender of names at any historical date (`gender`). In addition, we've started many packages of use in the archeology field, including a client for the Pleiades database (`pleiades`). Other packages are of use across social science disciplines, including for the Digital Public Library of America (`rdpla`), Europeana (`reuropeana`), and to scrape museum metadata (`musemeta`).

Last, we have many generic R clients for databases, including for Solr (`solr`), Elasticsearch (`elastic`), CKAN (`ckanr`), and ERDDAP (`rerddap`). These will be valuable to individual users, but will also facilitate more tools that use these databases, including in our own software.

- Community members committed code on `103` R packages in the last 15 months.
- `43` packages were updates of existing software,
- `60` packages were new software developed this year.
- `27` packages were new packages sent to the central R repository (CRAN) since the Sloan funding started.
- `8` packages were updated one or more times on CRAN.
- `83` packages are in active development.

## Outreach efforts

We spent countless many hours over the last year running workshops around the world.

**List of workshops**

- Oxford University
- University of Sheffield
- Open Knowledge foundation (Geneva)
- UC Berkeley (various workshops at D-lab, Berkeley Initiative in Global Change Biology, Geospatial innovation facility)
- National Center for Ecological Analysis and Synthesis, UC Santa Barbara
- Zoological Society of London
-  The Genome Analysis Center, Norwich Research Park, UK
-  The Commonwealth Scientific and Industrial Research Organisation (CSIRO), Canberra, Australia
-  The University of Melbourne
-  University of North Carolina
-  Duke University
-  University of British Columbia
-  Quebec Centre for Biodiversity Science
-  Science Online (Main event in North Carolina and Science Online Climate)

**Hackathon**

Besides such workshops, we also held a highly successful coding un-conference, the first of which we held in March 2014 in San Francisco ([http://ropensci.github.io/hackathon/](http://ropensci.github.io/hackathon/)).

## Engagement with our community

### Software developed

Participation in rOpenSci software development increased exponentially during the
grant period, both in number of contributors (Fig. 1) and number of code contributions
(Fig. 2). The number of package downloads (from one partial source, the
[RStudio mirror](http://cran-logs.rstudio.com/)) also increased x fold during this
time (Fig. 3).

```{r child="children/cumcontribs.Rmd", echo=FALSE, fig.cap='Cumulative number of contributors to rOpenSci packages from October 2011 to December 2014. The blue shaded area is the period of this grant.'}
```

```{r child="children/cumcommits.Rmd", echo=FALSE, fig.cap='Cumulative number of code commits to rOpenSci packages from October 2011 to December 2014. The blue shaded area is the period of this grant.'}
```

```{r child="children/crandownloads.Rmd", echo=FALSE, fig.cap='Cumulative downloads of rOpenSci packages from October 2011 to December 2014. Each line is a separate package.'}
```

### Engagement via the website

We reached our goal of 30,000 visitors for the grant period. In addition, via the chat client
on our site, our users have requested 157 conversations with us, with questions about software
installation or use. Last, we have received 144 conversations from users via our _Contact Us_ page.

### Impact across disciplines

We have engaged with two major disciplines in the social sciences: archeology and history.
We have talked with three leading digital archeologists, and two leading digital historians,
leading to a long list of potential future projects, and a few in development R packages
to connect archeologists and historians to open data on the web. We are making great progress
in these two fields and are starting to make connections in other social science fields,
including psychology. We are involved in a grant with an archeologist to work on R clients.

## rOpenSci citations and mentions in blogs/media

### List of Research Papers

* Miya M, Friedman M, Satoh TP, Takeshima H, Sado T, et al. (2013) Evolutionary Origin of the Scombridae (Tunas and Mackerels): Members of a Paleogene Adaptive Radiation with 14 Other Pelagic Fish Families. PLoS ONE 8(9): e73535. doi: 10.1371/journal.pone.0073535

* Gagnon, Yakir L., Sutton, Tracey T., Johnsen, Sönke. (2013) Visual acuity in pelagic fishes and mollusks. Vision Research 92: 1-9.

* Bartomeus, Ignasi, et al. (2013) "Biodiversity ensures plant–pollinator phenological synchrony against climate change." Ecology letters 16: 1331-1338.

* Goring, Simon, et al. (2013). "Pollen assemblage richness does not reflect regional plant species richness: a cautionary tale." Journal of Ecology 101: 1137-1145.

* Poisot, Timothée, Bérangère Péquin, and Dominique Gravel. (2013) "High Throughput Sequencing: A Roadmap Toward Community Ecology". Ecology and evolution 3: 1125-1139.

* Drozd, Pavel, and Jan Šipoš. (2013). "R for all (I): Introduction to the new age of biological analyses." Casopis slezskeho zemskeho muzea (A) 62: 51-57.

* Goodman, Alyssa, et al. (2014). "10 Simple Rules for the Care and Feeding of Scientific Data." arXiv preprint arXiv:1401.2134.

* Holman, L. et al. "Is there a preponderance of p values just under 0.05? Testing for p hacking across the biological sciences".

* Holman, L et al. "Evidence for strong and pervasive observer bias in biology: The importance of blind data recording".

* Price, S.A., et al. (2014). "Two waves of colonization straddling the K–Pg boundary formed the modern reef fish fauna". Proc. R. Soc. B vol. 281 no. 1783.

* Varela S, et al. (2014) "rAvis: An R-Package for Downloading Information Stored in Proyecto AVIS, a Citizen Science Bird Project". PLoS ONE 9(3)

* Talent, N., et al. (2014). "Character Selection During Interactive Taxonomic Identification:'Best Characters'." Biodiversity Informatics 9:1-12.

* Pandit, M.K., et al. (2014). "The contrasting effects of genome size, chromosome number and ploidy level on plant invasiveness: a global analysis." New Phytologist (In press).

* Altman, J., et al. "TRADER: A package for Tree Ring Analysis of Disturbance Events in R." Dendrochronologia 32(2) (2014): 107-112.

_(Other papers in progress are not reported here to protect confidentialty requested by some authors)_

__Papers by rOpenSci's team:__

* Chamberlain, S.A., and E. Szocs (2013). "taxize - taxonomic search and retrieval in R". F1000Research, 2:191. http://f1000research.com/articles/2-191/v2.

* Chamberlain, S.A. (2013). "Consuming Article-Level Metrics: Observations and Lessons from Comparing Aggregator Provider Data". Information Standards Quarterly 25(2): 5-13.

* F1000 special issue in preparation

*  RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R (in preparation by the team)

* One simple way to share your data (in preparation by the team). This paper is a response to the popular article _"Nine simple ways to make it easier to (re)use your data"_ published in Ideas in Ecology and Evolution.

### Websites

- [http://labs.biblioteca.uoc.edu/blog/?p=4332](http://labs.biblioteca.uoc.edu/blog/?p=4332) The Importance of Being Reproducible
- [http://www.infoworld.com/article/2608259/development-tools/github-rolls-out-the-red-carpet-for-scientists.html](http://www.infoworld.com/article/2608259/development-tools/github-rolls-out-the-red-carpet-for-scientists.html) GitHub rolls the red carpet for scientists
- [http://wiki.datadryad.org/External_Metadata_Use#ROpenSci](http://wiki.datadryad.org/External_Metadata_Use#ROpenSci)
- [http://paidcontent.org/2012/09/10/open-access-research-catastrophic-for-reed-elsevier/](http://paidcontent.org/2012/09/10/open-access-research-catastrophic-for-reed-elsevier/)
- [http://radar.oreilly.com/2013/01/kaitlin-thaney-open-science.html](http://radar.oreilly.com/2013/01/kaitlin-thaney-open-science.html)
- [http://sciencehackday.pbworks.com/w/page/24607895/Programming%20Tools%20and%20Frameworks](http://sciencehackday.pbworks.com/w/page/24607895/Programming%20Tools%20and%20Frameworks)
- [http://www.scilogs.com/mola_mola/big-data-for-big-ecology/](http://www.scilogs.com/mola_mola/big-data-for-big-ecology/)
- [http://captaincalliope.net/](http://captaincalliope.net/)
- [http://krr.cs.vu.nl/2012/11/trip-report-plos-article-level-metrics-workshop-and-hackathon/](http://krr.cs.vu.nl/2012/11/trip-report-plos-article-level-metrics-workshop-and-hackathon/)
- [http://postgenomic.com/blog/altmetric-in-2012-year-end-review/](http://postgenomic.com/blog/altmetric-in-2012-year-end-review/)
- [http://rossmounce.co.uk/2013/01/12/from-card-catalogs-to-computers-databases-in-vertebrate-paleontology/](http://rossmounce.co.uk/2013/01/12/from-card-catalogs-to-computers-databases-in-vertebrate-paleontology/)
- [http://ebooks.iospress.nl/publication/33465](http://ebooks.iospress.nl/publication/33465) (and paper in this folder: JahnEtal2013.pdf)
- [http://libreas.wordpress.com/category/libreas-visualisierung/](http://libreas.wordpress.com/category/libreas-visualisierung/)
