---
name: new-geospatial-stack
layout: post
title: The new geospatial stack
date: 2015-03-04
authors:
  - name: Scott Chamberlain
tags:
- R
- API
- geo
- spatial
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE
)
```

Geospatial data input/output, manipulation, and vizualization are tasks that are common to many disciplines. Thus, we're keenly interested in making great tools

## rOpenSci Geospatial tools

We have an increasing set of spatial tools for R:

### geojsonio 

[geojsonio](https://github.com/ropensci/geojsonio) - A tool for converting to and from geojson data. For example:

```{r}
library("geojsonio")
'xxx'
```

### wellknown 

[wellknown](https://github.com/ropensci/wellknown) - A tool for converting to and from well-known text data. For example:

```{r}
library("wellknown")
'xxx'
```

### cartographer

[cartographer](https://github.com/ropensci/cartographer) - xxdasfaff. For example:

```{r}
library("cartographer")
'xxx'
```

### gistr

[gistr](https://github.com/ropensci/gistr) - This is not a geospatial tool per se, but it's extremely useful for sharing maps. For example, with just a few lines, you can share an interactive map to GitHub. 

```{r}
library("gistr")
'xxx'
```

Additionally, we plan to help bring many geospatial analysis tools from Javascript by creating interfaces to them through R with the [V8](https://github.com/jeroenooms/V8/) package. One tool we'll build soon at our upcoming [unconference/hackathon](http://unconf.ropensci.org/) is an R client for [turf.js](http://turfjs.org/). Another we will builld is an R client for [proj4js](https://github.com/proj4js/proj4js), a Javascript library to do projections. Combining turf.js and proj4js will be make a nice set of tools. 

## Maps in R

### Static maps

* base
* [latticeExtra](http://cran.r-project.org/web/packages/latticeExtra/index.html)
* ggplot
* static versions of google maps, etc.

Data formats: 

Examples:

____

```{r}
library("maps")
library("latticeExtra")
data(USCancerRates)
suppressWarnings(print(
  mapplot(rownames(USCancerRates) ~ log(rate.male) + log(rate.female),
          data = USCancerRates,
          map = map("county", plot = FALSE, fill = TRUE,
                    projection = "mercator"))
))
```

### Interactive maps

* [htmlwidgets](http://www.htmlwidgets.org/)
* [leaflet](http://rstudio.github.io/leaflet/)
* [rMaps](https://github.com/ramnathv/rMaps)
* [cartographer](https://github.com/lmullen/cartographer)
* [plotly](https://github.com/ropensci/plotly)
* [leafletR](https://github.com/chgrl/leafletR)

Data formats: 
