---
name: ropensci-geospatial-stack
layout: post
title: Overview of the rOpenSci geospatial stack
date: 2016-03-10
authors:
  - name: Scott Chamberlain
tags:
- R
- API
- geo
- spatial
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE
)
```

Geospatial data input/output, manipulation, and vizualization are tasks that are common to many disciplines. Thus, we're keenly interested in making great tools. We have an increasing set of spatial tools.

## geojsonio

<a href="https://cran.rstudio.com/web/packages/geojsonio/"><span class="label label-warning">cran</span></a> <a href="https://github.com/ropensci/geojsonio"><span class="label label-info">github</span></a>

[geojsonio](https://github.com/ropensci/geojsonio) - A tool for converting to and from geojson data. For example:

```{r}
library("geojsonio")
'xxx'
```
## wellknown

<a href="https://cran.rstudio.com/web/packages/wellknown/"><span class="label label-warning">cran</span></a> <a href="https://github.com/ropensci/wellknown"><span class="label label-info">github</span></a>

[wellknown](https://github.com/ropensci/wellknown) - A tool for converting to and from well-known text data. For example:

```{r}
library("wellknown")
'xxx'
```

## gistr

<a href="https://cran.rstudio.com/web/packages/gistr/"><span class="label label-warning">cran</span></a> <a href="https://github.com/ropensci/gistr"><span class="label label-info">github</span></a>

[gistr](https://github.com/ropensci/gistr) - This is not a geospatial tool per se, but it's extremely useful for sharing maps. For example, with just a few lines, you can share an interactive map to GitHub.

```{r}
library("gistr")
'xxx'
```

## lawn

<a href="https://cran.rstudio.com/web/packages/lawn/"><span class="label label-warning">cran</span></a> <a href="https://github.com/ropensci/lawn"><span class="label label-info">github</span></a>

An R client for [turf.js](http://turfjs.org/).

```{r}
library("lawn")
'xxx'
```

## proj

<span class="label label-warning">cran</span> <a href="https://github.com/ropensci/proj"><span class="label label-info">github</span></a>

An R client for [proj4js](https://github.com/proj4js/proj4js), a Javascript library for projections.

## getlandsat

<span class="label label-warning">cran</span> <a href="https://github.com/ropenscilabs/getlandsat"><span class="label label-info">github</span></a>

## siftgeojson

<span class="label label-warning">cran</span> <a href="https://github.com/ropenscilabs/siftgeojson"><span class="label label-info">github</span></a>

## Maps

There's an increasing number of options for mapping in R. rOpenSci has two offerings in this space: `plotly` and `cartographer`.

### plotly

<a href="https://cran.rstudio.com/web/packages/plotly/"><span class="label label-warning">cran</span></a> <a href="https://github.com/ropensci/plotly"><span class="label label-info">github</span></a>

[plotly](https://github.com/ropensci/plotly) is an R client for [Plotly](https://plot.ly/) - a web interface and API for creating interactive graphics.

```{r}
library("plotly")
'xxx'
```

### cartographer

<span class="label label-inverse">cran</span> <a href="https://github.com/ropensci/cartographer"><span class="label label-info">github</span></a>

[cartographer](https://github.com/ropensci/cartographer) - xxdasfaff. For example:

```{r}
library("cartographer")
'xxx'
```

## Maptools Task View

<a href="https://github.com/ropensci/maptools"><span class="label label-info">github</span></a>

[Jeff Hollister](http://jwhollister.com/) is leading the [maptools task view](https://github.com/ropensci/maptools) to organize R mapping tools packages, sources of data, projections, static and interactive mapping, data transformation, and more.
