---
name: geospatial-suite
layout: post
title: The rOpenSci geospatial suite
date: 2016-11-16
authors:
  - name: Scott Chamberlain
  - name: Stefanie Butland
categories:
  - blog
tags:
- R
- geospatial
---


Since at rOpenSci we have an increasingly large set of tools for geospatial work in R, we thought it would be useful for us and readers to do an overview of these tools and future work. 

-----

First, we'll cover some vocabulary that we'll be using below.

* Well-known text (WKT)
  * links: [specification](https://d17oy1vhnax1f7.cloudfront.net/items/291d412o1c3E3s0R112A/06-103r4_Implementation_Specification_for_Geographic_Information_-_Simple_feature_access_-_Part_1_Common_Architecture_v1.2.1%20(1).pdf) - [Wikipedia entry](https://en.wikipedia.org/wiki/Well-known_text)
  * WKT is part of [Simple Features](https://en.wikipedia.org/wiki/Simple_Features) (see `sf` below)
  * Features of note: 
    * Character and binary representations
    * Supports any coordinate reference system
    * Often used in databases to store geospatial information
* GeoJSON
  * links: [specification](https://tools.ietf.org/html/rfc7946) - [Wikipedia entry](https://en.wikipedia.org/wiki/GeoJSON)
  * GeoJSON was inspired in part from Simple Features, but is not part of that specification. The most recent iteration is called [RFC 7946 GeoJSON](https://tools.ietf.org/html/rfc7946). 
  * Features of note: 
    * JSON character representation only (though see [geobuf](https://github.com/mapbox/geobuf) for binary GeoJSON - not part of RFC 7946)
    * All data is WGS84
    * Often found in web applications

Some features of WKT and GeoJSON overlap, while others do not.

Readers may be more familiar with shape files than WKT or GeoJSON. There are already great R tools for shape files, so our tools largely don't concern themselves with shape files and other types of geospatial files (e.g., KML, geotiff, etc.).

There are many geospatial tools in R already - e.g., check out the spatial task view, and check out our [maptools taskview](https://github.com/ropensci/maptools) for mapping tools and services in R. A noteable new tool in R geospatial is the package [sf](https://github.com/edzer/sfr) (simple features for R).

-----

## Use cases

In part to moivate our geospatial tools, we'll outline use cases for software we make that use WKT and GeoJSON. Thus, creating tools that fit our use cases makes a lot of sense. Some of our use cases:

* Web services that some of our packages interact with accept a geospatial filter as a query component. This often means WKT. Having tools that are light weight is important here as we don't need a full geospatial stack when we only need to lint WKT or create it from a bounding box, for example.
* Likewise, some web services only accept GeoJSON. Same argument as above follows here.
* Vizualize WKT and GeoJSON: given the above, users should be able to vizualize the area that they are defining with their WKT or GeoJSON. 
* WKT-GeoJSON conversion: sometimes users, and we, need to convert WKT to GeoJSON, or vice versa. Light weight tools to do that task are really useful.
* others?

-----

## The tools

rOpenSci has an increasing suite of database tools (with noteable recent releases):

### GeoJSON/WKT Tools

* [geoaxe](https://github.com/ropenscilabs/geoaxe) (split up well known text into chunks)
* [geojson](https://github.com/ropensci/geojson) (geojson classes for R) (on CRAN)
* [geojsonio](https://github.com/ropensci/geojsonio) (I/O for GeoJSON)
* [geojsonlint](https://github.com/ropenscilabs/geojsonlint) (Lint GeoJSON)
* [geoops](https://github.com/ropensci/geoops) (Operations on GeoJSON, sorta like `rgeos`) (on CRAN)
* [lawn](https://github.com/ropensci/lawn) (Turf.js R client)
* [wellknown](https://github.com/ropenscilabs/wellknown) (Well-Known-Text <--> GeoJSON) (on CRAN)


### Data/Data Services

* [geoparser](https://github.com/ropenscilabs/geoparser) (Geoparser.io client for place names)
* [rgeospatialquality](https://github.com/ropenscilabs/rgeospatialquality) (spatial quality of biodiversity records) (on CRAN)
* [getlandsat](https://github.com/ropenscilabs/getlandsat) (Landsat images) (on CRAN)
* [omsplotr](https://github.com/ropenscilabs/omsplotr) (OpenStreeMap data and vizualization) (on CRAN)
* [rnaturalearth](https://github.com/ropenscilabs/rnaturalearth) (Landsat images) (on CRAN)
* [geonames](https://github.com/ropensci/geonames) (Access geonames.org API) (on CRAN)

-----------

Let us know if you have any feedback on these packages, and/or if you think there's anything else we should be thinking about making in this space. Now on to some examples of our packages.

## Get devtools

We'll need `devtools` to install some of these packages, as not all are on CRAN. If you are on Windows, see [these notes](https://github.com/hadley/devtools#updating-to-the-latest-version-of-devtools).


```r
install.packages("devtools")
```

## pkg1

[pkg1](https://github.com/ropensci/xxxx) - pkg1 description

```{r eval = FALSE}
install.packages("pkg1")
```

short 1 par. description of what the pkg does, how it's useful, etc.

### Example

quick example ...

```{r eval = FALSE}
library("pkg1")
"xxx"
```

## pkg2

[pkg2](https://github.com/ropensci/xxxx) - pkg2 description

```{r eval = FALSE}
install.packages("pkg2")
```

short 1 par. description of what the pkg does, how it's useful, etc.

### Example

quick example ...

```{r eval = FALSE}
library("pkg2")
"xxx"
```

## pkg3

[pkg3](https://github.com/ropensci/xxxx) - pkg3 description

```{r eval = FALSE}
install.packages("pkg3")
```

short 1 par. description of what the pkg does, how it's useful, etc.

### Example

quick example ...

```{r eval = FALSE}
library("pkg3")
"xxx"
```

## etc. for prob. not all pkgs, but major ones

-----------

## What does all this mean for you?

Our goal with our geospatial tools is to make your work -- whether it be serious reproducible science, analysis of your company's data, or just fooling around with some data --  as easy as possible with as few installation headaches as possible.

xxxx??
