# https://datadryad.org/resource/doi:10.5061/dryad.63q27/3
# http://www.worldclim.org/current

This is an attempt to address this question from a Twitter conversation: https://twitter.com/corey_phillis/status/421731370029686784

> what i’d really like is to be able to cross-ref the dam db with rgbif. What fish spp. are above/below these dams? Corey Phillis (@corey_phillis)

To attack this problem we need dam data and GBIF data. We already have an R package for GBIF - we just need the dam data, hopefully retrievable in a programmatic fashion. 

## Get dam data

Download zip file and extract files

```{r getdata}
url <- 'http://sedac.ciesin.columbia.edu/downloads/data/grand-v1/grand-v1-dams-rev01/dams-rev01-global-shp.zip'
file <- basename(url)
download.file(url, file)
tmpdir <- tempdir()
untar(file, exdir = tmpdir)
list.files(tmpdir)
```

Rivers

```{r}
# url_rivers <- 'http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/50m/physical/ne_50m_rivers_lake_centerlines.zip'
# url_rivers <- 'http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/110m/physical/ne_110m_rivers_lake_centerlines.zip'
url_rivers <- 'http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/physical/ne_10m_rivers_lake_centerlines.zip'
file2 <- basename(url_rivers)
download.file(url_rivers, file2)
tmpdir2 <- tempdir()
untar(file2, exdir = tmpdir2)
list.files(tmpdir2)
```

Open shp files

```{r read}
library(rgdal)
dat <- readOGR(tmpdir, "GRanD_dams_v1_1")
head(dat@data[,1:5])
```

Get river data

```{r}
dat_rivers <- readOGR(tmpdir2, 'ne_10m_rivers_lake_centerlines')
head(dat_rivers@data)
plot(dat_rivers)
```

Get a single river

```{r}
dat_rivers@lines[[1]]
plot(SpatialLines(dat_rivers@lines[1:140]))
points(dat)
```


Plot to visualize dams

```{r plot}
plot(dat_rivers, col="blue")
points(dat)
```

Extract data

```{r extract}

```

## taxize - get some fish species

```{r taxize}
library(taxize)
# tsn <- get_tsn('Salmonidae')
# splist <- itis_downstream(tsn, downto='species')
splist <- c('Oncorhynchus mykiss', '')
```


## Get GBIF occurrence data

```{r gbif1, warning=FALSE, message=FALSE}
asdfsdf
```

## Visualize

```{r vis}
asdfsdf
```