---
title: Biodiversity comparisons across cities
pkg: rgbif
layout: usecases
---

In this example, we compare biodiversity in different regsions using GBIF data from rgbif.

This example can be done using BISON data as well with our rbison package.

```{r defaults, eval=TRUE, echo=FALSE, cache=TRUE}
opts_chunk$set(comment=NA, warning=FALSE, message=FALSE, fig.width=7, fig.height=6, fig.path="../assets/usecases-images/")
```

### Load libraries

```{r load}
library("rgbif")
library("ggplot2")
library("plyr")
```

### Get bounding boxes for some cites

```{r box}
box <- read.table("bounding.txt", sep="\t", col.names=c("city","minlat","maxlon","maxlat","minlon"))
```

### Get GBIF data using the rOpenSci package rgbif.

We'll not restrain our search to any particular taxonomic group, although you will likely do that in your own research. We'll make a new column with single lat/long coordinates for each cell for easy plotting. Last, we'll select 100 random cell IDs.

```{r getbounding}
temp <- gbifdata(densitylist(originisocountrycode="US"))
temp <- transform(temp, lat = (minLatitude+maxLatitude)/2, lon = (minLongitude+maxLongitude)/2)
cellids <- sample(temp$cellid, 100)
```

Then search for data for each of those cell ID's. We'll define a function to pass in each cell ID.

```{r getdata}
getdata <- function(x, maxresults=100){
  out <- occurrencelist(cellid=x, coordinatestatus=TRUE, maxresults=maxresults)
  df <- gbifdata(out, coordinatestatus=TRUE)
  data.frame(cellid=x, richness=length(unique(df$taxonName)))
}

out <- ldply(cellids, getdata)
out2 <- merge(out, temp[,c('cellid','lat','lon')], by='cellid')
# remove points outside the US
out3 <- out2[out2$lat < 49 & out2$lat > 24.7433195 & out2$lon > -130 & out2$lon < -66.9513812,]
```

### Plot data

```{r plot}
mapp <- map_data('state')
ggplot(mapp, aes(long, lat)) +
  geom_polygon(aes(group=group), fill="white", alpha=0, color="gray80", size=0.8) +
  geom_point(data=out3, aes(lon, lat, color=richness), size=3, alpha=0.8) +
  scale_color_gradient2(low = "white", mid="lightblue", high = "blue") +
  labs(x="", y="") +
  theme_grey(base_size=14) +
  theme(legend.position = "bottom", legend.key = element_blank())
```

### Notes

Bounding lat/long data from [here](https://raw.github.com/amyxzhang/boundingbox-cities/master/boundbox.txt)
